Temporal ordering and registration of cross-sectional imaging data

### 2014-05-14 image_analysis_paper/si_information.tex
started supplemental information for paper

%% PNAStmpl.tex
%% Template file to use for PNAS articles prepared in LaTeX
%% Version: Apr 14, 2008

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% BASIC CLASS FILE
%% PNAStwo for two column articles is called by default.
%% Uncomment PNASone for single column articles. One column class
%% and style files are available upon request from pnas@nas.edu.
%% (uncomment means get rid of the '%' in front of the command)

%\documentclass{pnasone}
\documentclass{pnastwo}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Changing position of text on physical page:
%% Since not all printers position
%% the printed page in the same place on the physical page,
%% you can change the position yourself here, if you need to:

% \advance\voffset -.5in % Minus dimension will raise the printed page on the
                         %  physical page; positive dimension will lower it.

%% You may set the dimension to the size that you need.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% OPTIONAL GRAPHICS STYLE FILE

%% Requires graphics style file (graphicx.sty), used for inserting
%% .eps files into LaTeX articles.
%% Note that inclusion of .eps files is for your reference only;
%% when submitting to PNAS please submit figures separately.

%% Type into the square brackets the name of the driver program
%% that you are using. If you don't know, try dvips, which is the
%% most common PC driver, or textures for the Mac. These are the options:

% [dvips], [xdvi], [dvipdf], [dvipdfm], [dvipdfmx], [pdftex], [dvipsone],
% [dviwindo], [emtex], [dviwin], [pctexps], [pctexwin], [pctexhp], [pctex32],
% [truetex], [tcidvi], [vtex], [oztex], [textures], [xetex]

%\usepackage[dvips]{graphicx}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% OPTIONAL POSTSCRIPT FONT FILES

%% PostScript font files: You may need to edit the PNASoneF.sty
%% or PNAStwoF.sty file to make the font names match those on your system.
%% Alternatively, you can leave the font style file commands commented out
%% and typeset your article using the default Computer Modern
%% fonts (recommended). If accepted, your article will be typeset
%% at PNAS using PostScript fonts.

% Choose PNASoneF for one column; PNAStwoF for two column:
%\usepackage{PNASoneF}
\usepackage{PNAStwoF}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% ADDITIONAL OPTIONAL STYLE FILES

%% The AMS math files are commonly used to gain access to useful features
%% like extended math fonts and math commands.

\usepackage{amssymb,amsfonts,amsmath}

%\usepackage{subcaption}
\graphicspath{ {paper_figures/} }

%\usepackage{sansmath}

%\usepackage[font={sf, small}]{caption}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% OPTIONAL MACRO FILES
%% Insert self-defined macros here.
%% \newcommand definitions are recommended; \def definitions are supported

%\newcommand{\mfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2}}
%\def\s{\sigma}

\DeclareMathSizes{9}{8}{7}{7}

\DeclareMathOperator*{\argmin}{arg\,min}

\makeatletter
\newcommand{\customlabel}[2]{%
\protected@write \@auxout {}{\string \newlabel {#1}{{#2}{}}}}
\makeatother

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Don't type in anything in the following section:
%%%%%%%%%%%%
%% For PNAS Only:
\contributor{Submitted to Proceedings
of the National Academy of Sciences of the United States of America}
\url{www.pnas.org/cgi/doi/10.1073/pnas.0709640104}
\copyrightyear{2008}
\issuedate{Issue Date}
\volume{Volume}
\issuenumber{Issue Number}
%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% For titles, only capitalize the first letter
%% \title{Almost sharp fronts for the surface quasi-geostrophic equation}

\title{Temporal ordering and registration of cross-sectional imaging data}

%% Enter authors via the \author command.
%% Use \affil to define affiliations.
%% (Leave no spaces between author name and \affil command)

%% Note that the \thanks{} command has been disabled in favor of
%% a generic, reserved space for PNAS publication footnotes.

%% \author{<author name>
%% \affil{<number>}{<Institution>}} One number for each institution.
%% The same number should be used for authors that
%% are affiliated with the same institution, after the first time
%% only the number is needed, ie, \affil{number}{text}, \affil{number}{}
%% Then, before last author ...
%% \and
%% \author{<author name>
%% \affil{<number>}{}}

%% For example, assuming Garcia and Sonnery are both affiliated with
%% Universidad de Murcia:
%% \author{Roberta Graff\affil{1}{University of Cambridge, Cambridge,
%% United Kingdom},
%% Javier de Ruiz Garcia\affil{2}{Universidad de Murcia, Bioquimica y Biologia
%% Molecular, Murcia, Spain}, \and Franklin Sonnery\affil{2}{}}

\author{Carmeline~J.~Dsilva\affil{1}{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA},
Bomyi~Lim\affil{1}{},
Thomas~J.~Levario\affil{2}{School of Chemical and Biomolecular Engineering, Georgia Institute of Technology, Atlanta, Georgia, USA},
Hang~Lu\affil{2}{},
Amit~Singer\affil{3}{Department of Mathematics, Princeton University, Princeton, New Jersey, USA} \affil{4}{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA},
Stanislav~Y.~Shvartsman\affil{1}{} \affil{5}{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA},
\and
Ioannis~G.~Kevrekidis\affil{1}{} \affil{4}{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA}}

\contributor{Submitted to Proceedings of the National Academy of Sciences
of the United States of America}

%% The \maketitle command is necessary to build the title page.
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{article}

\begin{abstract}
In studies of development, researchers are often presented with cross-sectional data, where each data point is a sample from a population fixed at a slightly different developmental time.
%
The goal is then to temporally order the data to reconstruct the developmental dynamics.
%
If each data point is a two-dimensional image, the images must first be registered before they can be temporally ordered.
%
When such data sets are large, noisy, and/or if the developmental changes are subtle, these tasks can be difficult to do by hand.
%
We present an automatic approach to register {\it and} temporally order cross-sectional data sets of images.
%
The mathematical techniques (angular synchronization for image registration, diffusion maps for temporal ordering, and
vector diffusion maps for simultaneously performing both tasks) are applicable to a wide variety of data sets and
require little {\it a priori} knowledge of the image features or the developmental dynamics.
%
We demonstrate the utility of our methods using a collection of images from a study of {\it Drosophila} embryogenesis.
\end{abstract}

%% When adding keywords, separate each term with a straight line: |
\keywords{temporal ordering | image registration}

%% Optional for entering abbreviations, separate the abbreviation from
%% its definition with a comma, separate each pair with a semicolon:
%% for example:
%% \abbreviations{SAM, self-assembled monolayer; OTS,
%% octadecyltrichlorosilane}

% \abbreviations{}

%% The first letter of the article should be drop cap: \dropcap{}
%\dropcap{I}n this article we study the evolution of ''almost-sharp'' fronts

%% Enter the text of your article beginning here and ending before
%% \begin{acknowledgements}
%% Section head commands for your reference:
%% \section{}
%% \subsection{}
%% \subsubsection{}

\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal and cross-sectional \cite{diggle2002analysis}.
%
In longitudinal studies, developmental progress is monitored over time for the same embryo.
%
In a cross-sectional study, one embryo contributes only a single snapshot of a chemical or morphological process along its developmental trajectory, and the developmental dynamics must be reconstructed from multiple snapshots of different embryos.
%
Both of these sampling schemes have their advantages and limitations, and both are extensively used by developmental biologists.
%
Here we focus on cross-sectional studies, which have a time-honored history and still present the only option for most organisms.
%
In a typical cross-sectional study, a group of developing embryos is fixed using a procedure that arrests their development and stained with chemicals that help visualize a handful of cellular processes.
%
Fixed embryos are then imaged using any given number of microscopy techniques.
%
Recent advances in large-scale physical manipulation and imaging of embryos have produced rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different geometric orientation and developmental time point.
%
Importantly, the ``age'' of any given embryo arrested in its development is not known to high accuracy.
%
In general, it is only known that a collection of embryos belongs to a certain time window.
%
In order to recover the developmental dynamics from such image datasets, snapshots of different embryos must be spatially aligned or registered to factor out the relevant symmetries (i.e., translations and rotations), and then ordered in time.
%
We present a general computational framework which greatly accelerates -and can combine- both of these tasks.

Temporal ordering and registration of images is straightforward when the number of images is small and differences between them are visually apparent.
%
As an example, Figure~\ref{fig:fish} shows a caricature of fish development, combining the processes of growth and color patterning.
%
In this case, temporal ordering can be accomplished by arranging the fish by size, which is monotonic with the developmental progress.
%
Image registration is based on obvious morphological landmarks, such as the position of head and fins.
%
On the other hand, real data poses nontrivial challenges, due to the number of images, measurement noise and variability, and the subtlety of the developmental changes.

To illustrate the issues associated with temporal ordering and registration of images, we use a data set from an imaging study of signaling protein expression in the early fruit fly embryo, one of the leading model systems for studies of developmental dynamics.
%
This dataset comprises $49$ fluorescence microscopy images of different embryos fixed during the $3^{rd}$ hour of their development.
%
\begin{figure}[t]
\includegraphics[width=8.4cm]{fig1}
\caption{{\it (A)} Fish, each in a different orientation and a different stage of development. {\it (B)} Fish, now rotationally registered and temporally ordered. For this caricature, the registration and ordering is easy to do ``by hand" because the data set is small and the developmental changes are easy to recognize.}
%\label{fig:fish}
\customlabel{fig:fish}{1}
\customlabel{subfig:fish_unordered}{\ref{fig:fish}{\it A}}
\customlabel{subfig:fish_ordered}{\ref{fig:fish}{\it B}}
\end{figure}
%
The fruit fly embryo is shaped like a rice grain, approximately $1/2$~mm long.
%
By end of the $2^{nd}$ hour of development, sequential divisions of a fertilized egg generate a system where $\sim 6,000$ nuclei are uniformly arranged under the common plasma membrane.
%
During the $3^{rd}$ hour, the embryo undergoes the process of cellularization, whereby membranes that enclose nuclei into different cells grow in thickness (Figure~\ref{subfig:membranes}).
%
In conjunction with these morphological changes, concentration profiles of multiple regulatory molecules form inside the embryo, subdividing it into regions that give rise to future tissues and organs \cite{lim2013kinetics}.
%
Images in Figure~\ref{subfig:images_unordered} show optical sections at a fixed depth, perpendicular to the long axis of the embryo.
%%%
%%% we have to somehow say that it is consistent
%%%
%
%%%
%%% ask if "visualize" is typical as active verb
%%%
The embryos were fixed and stained with chemicals that help visualize nuclei (shown in gray) as well as the spatial distributions of two different proteins (shown in green and red) \cite{chung2010microfluidic}.
%
Each of the images reveals a uniform arrangement of nuclei and a nonuniform distribution of the two proteins, which are involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig2}
\caption{{\it (A)} Fluorescent images of {\it Drosophila} embryos. Each embryo has been stained for Dorsal (green) and dpERK (red). {\it (B)} Selected image from data set. {\it (C)} Profile of Dorsal (green) and dpERK (red) around the circumference of the embryo in {\it B}. }
%\label{fig:images_unordered}
\customlabel{fig:images_unordered}{2}
\customlabel{subfig:images_unordered}{\ref{fig:images_unordered}{\it A}}
\customlabel{subfig:one_image}{\ref{fig:images_unordered}{\it B}}
\customlabel{subfig:image_profiles}{\ref{fig:images_unordered}{\it C}}
\end{figure}

%%%
%%% orientation does not read good to Y below -
%%% it is the same orientation, but a different rotation around the axis
%%% how to say this ?
%%%
Each image in Figure~\ref{subfig:images_unordered} comes from a different embryo, captured at a different rotation about the long axis and at a different and unknown point in time.
%
The goal is to register these images in a globally consistent way, and order them to be representative of the underlying developmental trajectory.
%
Both of these tasks can be fully automated, and even combined.
%
We will use {\it angular synchronization} \cite{singer2011angular}, a registration algorithm that is robust to noise, to align the images in a consistent way.
%
We will use {\it diffusion maps} \cite{coifman2005geometric}, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets, to temporally order our data.
%
We will also show how both steps can be combined using {\it vector diffusion maps} \cite{singer2012vector}.
%
%The paper is organized as follows.
%
%First, we use a synthetic dataset, where the correct orientation and ordering of data points is known, to formalize the tasks of image registration and temporal ordering.
%
%Then, we show that both of these tasks can be effectively combined in a single step.
%
%Following this, we return to the original set of two-dimensional images and demonstrate that our approach can readily align them and order them correctly in time.
%
%We conclude by discussing the general requirements that must be satisfied for our approach to work and compare it to other approaches and ideas in cross-sectional studies of biological dynamics.

\section{Results and Discussion}

%We would like to temporally order images, such as the images shown in Figure~\ref{subfig:images_unordered}.
%
%We will use diffusion maps, a dimensionality reduction algorithm, to automatically order our data.
%
%However, this algorithm requires distances between data points.
%
%Because each image is in a different orientation, we need to first align or {\it register} the images before computing distances.
%
%Once all the images have been aligned, we can compute distances and use diffusion maps to temporally order the data.

%%%
%%%  Y to discuss with S
%%%  I think we should have the real data
%%%
%%%  also, these dynamics "are like" our dynamics
%%%
%%%   even if this is kept, one should say that this chosen because it LOOKS LIKE what we have
%%% another reason I want to have the real data is because in the discussion I want to talk about the flips
%%%

We will first demonstrate our techniques for registration and temporal ordering using a synthetic data set whose features resemble the image features in Figure~\ref{fig:images_unordered}.
%
The relatively simple dynamics of this data set will allow us to easily visualize and illustrate the main points of the different algorithms.
%
We construct concentration profiles defined on a ring (as in Figure~\ref{subfig:one_image}), and we assume that each ring is arbitrarily rotated around its center; an example is shown in Figure~\ref{subfig:1d_example}.
%
Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown at the bottom of Figure~\ref{subfig:1d_example} (the symmetry group is $SO(2)$, the group of all two-dimensional proper rotations).
%
{\bf We should discuss/mention that the eggs could be flipped.}
%
Each concentration profile is a noisy Gaussian (shown in Figure~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
%
We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors (the corresponding symmetry group for the discretized profiles is $\mathbb{Z}_{100}$, the group of integers modulo $100$).
%
Figure \ref{subfig:1d_unaligned_unordered} shows the entire data set; the concentration profiles have been stacked in an array, so that each row corresponds to a single profile.
%
Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian whose amplitude grows in time) are not readily apparent.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig3}
\caption{{\it (A)} One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom). {\it (B)} Intensity corresponding to the profile in {\it A}. {\it (C)} An ensemble of concentration profiles, each of the form described in {\it A}. Each row in the array corresponds to a single profile. {\it (D)} The profiles in {\it C}, now registered using angular synchronization. {\it (E)} The profiles in {\it D}, now temporally ordered using diffusion maps. {\it (F)} The profiles in {\it C}, registered and temporally ordered in a single step using vector diffusion maps.}
%\label{fig:1d_demo}
\customlabel{fig:1d_demo}{3}
\customlabel{subfig:1d_example}{\ref{fig:1d_demo}{\it A}}
\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}{\it B}}
\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}{\it C}}
\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}{\it D}}
\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}{\it E}}
\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}{\it F}}
\end{figure}

\subsection{Angular synchronization for image registration}

We begin by registering the set of profiles using angular synchronization\cite{singer2011angular}, a technique which is robust to noise and requires no {\it a priori} knowledge of image features.
%
Angular synchronization first aligns every possible pair of data points, i.e., it computes the rotation angles
which make each pair have the maximum possible overlap (see {\it Materials and Methods}).
%%%
%%% so I do not like the figure with the vectors - it is correct, but people have to take the
%%% function, discretize and make it a vector in their heads, and may be "one step too much"
%%%
%
An illustration is shown in Figure~\ref{subfig:synch1};
each data point is a vector, and $\theta_{ij}$ denotes the angle of rotation needed to align vectors $i$ and $j$.
%
{\bf It is not clear how we are rotating the arrows (around the center vs. around the tail).}
%
From this large collection of pairwise alignments, the algorithm then selects a set of rotation angles, $\theta_i$ (one for each data point), such that differences in rotation angles, $\theta_i-\theta_j$, are {\it most consistent} (in some metric) with the pairwise alignments $\theta_{ij}$.
%
Figure~\ref{subfig:synch2} shows each vector from Figure~\ref{subfig:synch1}, now rotated so that the entire set is aligned; the computed rotation angles are also indicated.
%
Note that the differences between the rotation angles assigned to each vector are consistent with the pairwise angles needed to rotate one vector to another, i.e., $\theta_i - \theta_j \approx \theta_{ij}$.
%
The problem of finding these alignments is formulated as an optimization problem, and then relaxed to an eigenproblem.
%
The top eigenvector(s) of a matrix (which contains information about all pairwise alignments)
provides the rotation angles (up to a constant rotation) assigned to each data point (see {\it Materials and Methods}).
%
%Because the method considers {\it all} pairwise alignments, angular synchronization is often robust to noise.
%
%
Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles from Figure~\ref{subfig:1d_unaligned_unordered}, now aligned using angular synchronization.
%
It is reassuring that the peaks appear aligned,
even though the algorithm uses no specific information about the existence or location of such peaks.
%
%We note once more that, since angular synchronization only uses pairwise rotation information,
%the optimal rotations are recovered up to a global rotation.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig4}
\caption{ {\it (A)} Set of vectors, each in a different orientation. The pariwise alignment angles are indicated. {\it (B)} The vectors from {\it A}, each rotated so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in {\it A}. {\it (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of data points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see \eqref{eq:dmaps_W}), so that close data points are connected by dark (``stronger'') edges. {\it (D)} The data in {\it C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure, $\phi_2$. The color intensity is monotonic with the arclength, thus parameterizing the curve.}
%\label{fig:schematics}
\customlabel{fig:schematics}{4}
\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
\customlabel{subfig:dmaps_edges}{\ref{fig:schematics}{\it C}}
\customlabel{subfig:dmaps_color}{\ref{fig:schematics}{\it D}}
\end{figure}

\subsection{Diffusion maps for temporal ordering}

After geometrically registering the data (concentration profiles or images), we want to order them in time to reconstruct the developmental dynamics.
%
%Our main assumption is that most of the variation in the data correlates with time.
%
Our main assumptions are that the high-dimensional data (one dimension per discretization point in the concentration profiles/one dimension per color pixel intensity in the images) lie on a one-dimensional, nonlinear curve, {\it and} that this curve is parameterized by time.
%
Figure~\ref{subfig:dmaps_edges} shows a schematic of a two-dimensional data set (the ``high-dimensional" data) which lies on a one-dimensional nonlinear curve (the ``low-dimensional" manifold).
%
We assume that the data points are ordered from youngest to oldest along the curve.
%
When our data are very high dimensional, we cannot extract this curve by eye.
%
We instead use diffusion maps \cite{coifman2005geometric}, a dimensionality reduction technique which aims to uncover low-dimensional, nonlinear structure in high-dimensional data sets, to automatically extract this one-dimensional curve.
%
Figure~\ref{subfig:dmaps_color} shows the same data, now colored by the leading (non-trivial) diffusion map coordinate.
%
The color parameterizes the one-dimensional curve that our eye can visually detect, and ordering the data by the leading diffusion map coordinate will effectively order data in time.

To uncover the diffusion map parameterization, we assume that data points which are close in the high-dimensional space should also be close {\it on the curve} (we would like images which look similar to be close together in our developmental trajectory).
%
%We therefore want to find a parameterization of the points, such that the points which are close in the original, high-dimensional space are {\it as close as possible} in the parameterization.
%
To emphasize this point, edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored based on the distance between the data points (using the diffusion kernel of the distance), so that pairs of points which are close have dark/strong edges connecting them.
%
Visually, the dark edges delineate the one-dimensional curve, whereas the light edges are less informative about the low-dimensional structure.
%
%We want points which are close in the high-dimensional space, i.e., connected by dark edges, to be close in our low-dimensional %parameterization.
%
Finding such a parameterization can be formulated as an optimization problem \cite{Belkin2003}, and the solution is given by the top (non-trivial) eigenvector of a Markov matrix which contains information about pairwise distances between the data points (see {\it Materials and Methods}).
%
%This procedure has been named diffusion maps \cite{coifman2005geometric}.
%
%The formulation extends from curves to manifolds (i.e., we can search for an $n$-dimensional, rather than a one-dimensional, parameterization of the data, such that data points which are close in the high-dimensional space are also close in the $n$-dimensional parameterization).

Given our set of registered signals, shown in Figure~\ref{subfig:1d_aligned_unordered}, we use diffusion maps to order them in time.
%
We use the Euclidean distance between the (discretized in $\mathbb{R}^{100}$) aligned concentration profiles (the rows of Figure~\ref{subfig:1d_aligned_unordered}) as our pairwise distances between data points.
%
Figure~\ref{subfig:1d_aligned_ordered} shows the data from Figure~\ref{subfig:1d_aligned_unordered}, now sorted by the value of the leading diffusion map coordinate, $\phi_2$.
%
As expected, the intensity of the peak grows as a function of $\phi_2$ (since we constructed these data
to have a peak growing in time).
%
%Although diffusion maps can order the data, it cannot tell us about the speed of this trajectory.
%
%Furthermore, the direction of the trajectory is also unknown; determining which end of the trajectory corresponds to the beginning of development is a task that needs to be done by hand.

\subsection{Vector diffusion maps for simultaneous registration and ordering}

%We are often interested in both registering and ordering our data.
%
%Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
%
Angular synchronization (for registration) can be combined with diffusion maps (for temporal ordering) in a single eigencomputation that allows us to {\it simultaneously} recover the registrations {\it and} the temporal ordering.
%
This technique is called vector diffusion maps \cite{singer2012vector} (see {\it Materials and Methods}).
%
It both reduces the number of required eigencomputations, and is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
%
Figure~\ref{subfig:1d_aligned_ordered_vdm} shows the result of simultaneously registering and ordering the data in Figure~\ref{subfig:1d_unaligned_unordered} using vector diffusion maps.
%
The data has been ordered by $\psi_{1, 3}$, the leading vector diffusion map coordinate.

\subsection{Analysis of two-dimensional images}

The same methods (with small modifications) can be applied directly to our two-dimensional images.
%
Instead of one-dimensional profiles (discretized to $100$-dimensional concentration vectors), we now work with
two color, two-dimensional images (discretized into $100 \times 100$ pixels/$20000$-dimensional vectors), and we register the images with respect to translations and rotations (the symmetry group is $ISO(2)$).
%
To facilitate the application of vector diffusion maps, we approximate the two-dimensional translations and rotations by {\it three-dimensional rotations} (symmetry group $SO(3)$; see {\it SI Appendix}).
%
We use vector diffusion maps to align and order the images shown in Figure~\ref{subfig:images_unordered};
the results are shown in Figure~\ref{fig:images_ordered}.
%
Both the red and green peaks are now aligned, and the intensity of the red peaks grows as a function of the leading vector diffusion map coordinate.

In these experiments, we are fortunate enough to be able to validate our  automatic registration and ordering using prior physical knowledge about the system.
%
For registration, it is known that the peak in the distribution of Dorsal (shown in green) specifies the ventralmost position in the embryo, and that the nuclei (shown in gray) delineate the circumference of the embryo.
%%%
%%% maybe you can also quantify this guy
%%%
%
%Based on this, images can be translated to make the centers of the embryos coincide and rotated to place the peak of the green channel at the same angular position.
%
Temporal ordering, on the other hand, can be done based on the monotonic progress of cellularization (Figure~\ref{subfig:membranes}) \cite{figard2013plasma}.%; this helps validate our vector diffusion map-based ordering.
%
The two methods for registration and ordering can be visually compared through the sequences of one-dimensional profiles extracted from the circumference of the embryos (similar to the format in Figure~\ref{fig:1d_demo}), shown in Figures~\ref{subfig:1d_membranes}~and~\ref{subfig:1d_vdm}.
%
The general trends are consistent:
both green  and red peaks are aligned, and the red peaks grow as a function of time/leading vector diffusion map coordinate.
%
However, Figure~\ref{subfig:1d_vdm}, which was automatically registered and ordered, shows an arguably better clustering of the ``outer" peaks in the top corners of the image.
%
%These peaks are more scrambled in Figure~\ref{fig:membrane_compare}b, which was analyzed using cellularization.
%
These peaks are known to arise later in development, and so it is reassuring that they appear at the end of the (recovered) trajectory.
%
Furthermore, the membranes stop growing towards the end of this trajectory, making the ordering of the late signals in Figure~\ref{subfig:1d_membranes} less accurate.
%
Quantitatively, the rank correlation coefficient between these two temporal orderings is 0.9158, indicating that our automatic ordering is consistent with the ordering obtained from cellularization.

%Although our methods are fairly robust, they do require a sufficient amount of data so that DMAPS can ``see'' a smooth trajectory.
%
%We evaluated how much data is required by subsampling our data set (with replacement), and for each subsample, computing the temporal ordering using VDM and calculating the rank correlation between this ordering and the ordering from cellularization.
%
%The results are shown in Figure~\ref{subfig:bootstrap} (each data point is averaged over 50 random subsamples).
%
%For small data sets, the average rank correlation is very low, indicating that the orderings from VDM are inaccurate.
%
%However, for moderately sized data sets (approximately 30 data points), the rank correlation begins to plateau and VDM can accurately order the data.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig5}
\caption{The images in Figure~\ref{subfig:images_unordered}, aligned and temporally ordered using vector diffusion maps. The images are sorted according to the value of the first non-trivial vector diffusion map component, $\psi_{1, 4}$.}
%\label{fig:images_ordered}
\customlabel{fig:images_ordered}{5}
\end{figure}

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig6}
\caption{{\it (A)} Images, stained for membranes (gray) and Dorsal (green) at different levels of development. {\it (B)} Membrane length as a function of time. The points corresponding to the images in {\it A} are indicated with arrows. {\it (C)} The one-dimensional concentration profiles, registered by peak Dorsal fluorescence and ordered by membrane length. {\it (D)} The one-dimensional concentration profiles corresponding to the automatically registered and ordered images in Figure~\ref{fig:images_ordered}. }
%\label{fig:membrane_compare}
\customlabel{fig:membrane_compare}{6}
\customlabel{subfig:membranes}{\ref{fig:membrane_compare}{\it A}}
\customlabel{subfig:membrane_curve}{\ref{fig:membrane_compare}{\it B}}
\customlabel{subfig:1d_membranes}{\ref{fig:membrane_compare}{\it C}}
\customlabel{subfig:1d_vdm}{\ref{fig:membrane_compare}{\it D}}
\end{figure}

\section{Conclusions}

We have shown how symmetry and dimensionality reduction techniques can successfully register and order cross-sectional developmental imaging data.
%
Although we illustrated our methods using a specific data set from a study of {\it Drosophila} embryogenesis, the framework is quite general and requires little knowledge of the image features or the developmental dynamics.
%
To our knowledge, temporal ordering of {\it Drosophila} images has not been automated to this extent.
%
Recently, machine learning algorithms have been used to group {\it Drosophila} images into different developmental classes \cite{yuan2014automated}, yet there, the appropriate classes must be defined by the user and hand-labeled training data are required.

Temporal ordering of cross-sectional biological data has been studied in a variety of other contexts.
%
Cross-sectional RNA sequencing data has been temporally ordered \cite{anavy2014blind, trapnell2014dynamics}, where each data point is the transcriptome for a single sample.
%
Gene expression data from multiple patients have been studied in \cite{gupta2008extracting, qiu2011discovering};
here, each data point is a vector containing the expression levels of the various genes.
%
Snapshots of cells throughout the cell cycle can be described by a vector of features (e.g., amount of DNA) which quantify the cell's state  \cite{kafri2013dynamics}.
%
These problems are no different than our task of ordering images, each of which is a large vector of pixel intensities.
%
The majority of these works temporally order the data either by solving a traveling salesman problem on the data, or by constructing a minimum spanning tree of the data,
expecting that these structures (the path of the traveling salesman, or the minimum spanning tree) characterize the majority of the structure/variability within the data and provide an accurate encoding of the dynamical progression.
%
We, instead, use diffusion maps to construct a one-dimensional parameterization of the data.
%
Diffusion maps are one of many nonlinear dimensionality reduction techniques that have been recently developed \cite{Belkin2003, tenenbaum2000global, Donoho2003, Roweis2000}.
%
They have been shown to be more robust to noise than other path-based algorithms, such as isomap \cite{balasubramanian2002isomap}, and so we suspect they may perform better than the ordering algorithms used in previous work.
%
{\bf Yannis says ``careful'', he thinks this might be too strong of a statement.}
%%%
%%%   we need help here
%%%
%
%Furthermore, although we use only the first diffusion maps coordinate to order the data (we assume the data are one-dimensional), diffusion maps are not limited to one-dimensional data and can be used to extract more complex structures.
%
%For example, in data which contains several branches or classes, the first few coordinates will delineate the branches or separate the classes.

The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
%
The main point to note here is that many registration algorithms require definition of appropriate landmarks or features for the images and the location of these within each image.
%
For example, in face recognition, the first step is often to find the eyes, nose, and mouth within the face, and then register images by aligning these features \cite{zhao2003face}.
%
Temporally ordering images of a construction site over time is also based on first identifying features within the images \cite{schindler2007inferring}.
%
In general, appropriate features may not be known {\it a priori}, and so we require a registration algorithm that does not rely on predefined features, but only on the inherent geometric symmetry of the problem.
%
Template-based methods are also common for registration, where one optimally aligns each image to a predefined template function.
%
However, these methods can suffer when the images are noisy, and the success of such methods can depend strongly on the template function \cite{shatsky2009method}.
%
{\bf Need to discuss consistency.}

Angular synchronization and vector diffusion maps require no feature points and no template function.
%
These methods have been used to register two-dimensional cryo-electron microscopy images of three-dimensional molecules \cite{singer2011three}, where each image is a picture of (essentially) the {\it same} molecule; variation in the data comes from differences in the viewing angles (rotations of the molecule) and from instrument/measurement noise.
%
In contrast, our data contains variation from the different embryo orientations, as well as variation from the underlying developmental dynamics. % (which underpins the development of vector diffusion map algorithms).
%
In addition to the eigenproblem relaxation used here, other synchronization algorithms have been proposed \cite{bandeira2014multireference}.
%
Synchronization is formulated for general symmetry groups, and depending on the data, one might consider other symmetries, such as reflections and scaling/dilation.
%
Here, the main requirement is that the symmetry group be represented by orthogonal or unitary matrices \cite{singer2013spectral}.

Temporal ordering can also be accomplished without first registering the images, by constructing functions of the images which are invariant under translations and rotations. 
%
These functions are then used as the data in the diffusion maps calculation; since these representative functions do not change under translations and rotations, the resulting diffusion map embedding is invariant to symmetries.
%
Such functions include the scale invariant feature transform \cite{lowe1999object, lowe2004distinctive}, the scattering transform \cite{mallat2012group}, and the bispectrum \cite{zhao2014rotationally}.
%
These methods have the disadvantage that we do not obtain a registered set of images, but only a temporal ordering.
%
However, such methods can often be faster for large data sets, since the functions are computed per image, rather than computing pairwise quantities in synchronization.

>>>>

\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}

\usepackage{graphicx}
\graphicspath{ {paper_figures2/} }

\usepackage{amssymb,amsfonts,amsmath}
\DeclareMathOperator*{\argmin}{arg\,min}

@@

\section{Experimental Methods}

\subsection{Measuring developmental time during cellularization}

\begin{figure}[h!]
\includegraphics[width=8.4cm, trim=0cm 5.6cm 0cm 0cm, clip]{fig6}
\caption{{\it (A)} Images, stained for membranes (gray) and Dorsal (green) at different levels of development. {\it (B)} Membrane length as a function of time. The points corresponding to the images in {\it A} are indicated with arrows.}
\label{fig:membrane_compare}
\end{figure}

\section{PCA for gastrulating images}

\begin{figure}[h!]
\raisebox{5.2cm}{A}
\includegraphics[width=16.8cm]{data2_PCA_ordered}

\raisebox{3.7cm}{B}
\includegraphics[width=8.4cm]{data2_PCA_variance}
\caption{{\it A} Images shown in Figure~4 of the paper, now ordered by the first PCA projection coefficient. Note that the ordering is not consistent with the developmental dynamics. {\it B} The eigenvalue spectrum from PCA. Note that subsequent components have a nonnegligible contribution on the data.}
\end{figure}

\section{Mathematical Algorithms}

\subsection{Angular synchronization \cite{singer2011angular}}
